Statistical quantification of confounding bias in predictive modelling

### 2021-10-30 manuscript.tex
finalized for preprint

Estimates of statistical power for the partial confounder test (with normal and linear simulations, for  a wide range of parameters) are shown on Figure \ref{fig:sim-normal}. Notably, with sample sizes as large as 1000, a confounder contributing only $\sim 4\%$ to the variance of the predictions can already be robustly detected ($w_{c\hat{y}} = 0.2$) with a power of 94-100\%. With a sample size of 500, the same confounding bias is still detected with a power greater than 84-100\% in all of the simulation cases. A sample size of 100 requires a somewhat stronger bias with approximately 12\% of explained variance ($w_{c\hat{y}}=0.4$) to achieve a reasonable level of power (75-98\%). Finally, even with a relatively low sample size of 50, the same amount of confounder variance is detected with a power of at least 50\%. If the confounder explains more than 25\% of variance, it is almost certainly detected even with a low sample size of $n \geq 50$.

Simulations show that non-normality has minimal effect on the power of the tests, except in case of extreme non-normality. (Fig. \ref{fig:sim-non-normal}). Simulations with sigmoid dependence resulted in an apparent loss of statistical power, however this is simply a consequence of the simulation methodology: with the same parameters, the sigmoid transformed confounder explains only approximately half the variance as compared to linear simulations.
Type I error control was valid in case of categorical variables and for the \emph{full} confounder test, as well. Power in these cases was also highly similar to the partial confounder test results with numerical variables. (Supplementary Figures \ref{fig:sim-bbb-lin-partial}, \ref{fig:sim-bbb-lin-full}, \ref{fig:sim-bbb-sig-partial} and \ref{fig:sim-bbb-sig-full}).

@@

Both acquisition batch and age group were statistically significantly associated with fluid intelligence ($R^2=0.032$ and $0.011$ and $p<0.001$ and $p=0.001$, respectively, see also Table \ref{tab:unconditional-pvals}). The model trained on the raw (unadjusted) connectivity features predicted fluid intelligence with a medium effect size ($R^2=0.095$, $p<0.001$).

@@

The full confounder test was highly significant ($p<0.0001$) for all models, indicating that neither of the models were exclusively driven by the confounds.

@@

The full confounder test was strongly significant ($p<0.0001$) for all models, except when regressing-out motion from the features ($p=0.09$), indicating that this model might have been (almost) exclusively driven by motion artifacts.

@@

\end{tabular}
\caption{\label{tab:unconditional-pvals} Coefficients-of-determination ($R^2$), the corresponding p-values and the p-values of the partial and full confounder tests, for all investigated confounders (conf.) confounder-mitigation methods (method). Bold numbers denote significant confounding bias identified by the partial confounder test or the lack of significant contribution over that of the confounder, as identified by the full confounder test.}
\end{table}

@@

The assessment of confounding bias via conditional independence testing is a very appealing solution due to its simplicity and applicability in a wide variety of settings. However, statistical testing of conditional independence is challenging: 'conditional analogs' of bivariate (unconditioned) non-parametric tests can give invalid results even by slight violations of normality and linearity\citep{korn1984ranges} (see an example on Fig. \ref{fig:sim-h0-demo}). While the magnitude of this problem may not be fully appreciated in case of predictive model diagnostics, the "no free lunch theorem" (i.e. the nonexistence of a uniformly valid conditional independence test)\cite{shah2020hardness} clearly implies that conditional independence-based confounding bias testing must be designed so that its suitability for the particular problem may be judged easily.

@@

As expected from theory, both tests displayed a valid type I error control and a high, practically relevant statistical power in the simulations, even if both the predictions and the confounder are non-normally and/or non-linearly dependent on the target variable (except extreme non-normality), confirming that the tests can be deployed in a wide variety of predictive modelling scenarios. While different biomedical applications might consider different amounts of bias to be relevant, the presented results can serve as a basis for power calculation in order to identify the necessary sample size for proper model diagnostics.

A characteristic example for the potential areas of applications is the novel field of "predictive neuroscience", where applying predictive modelling and machine learning on functional neuroimaging data holds great potential for both revolutionizing our understanding of the physical basis of mind and delivering clinically useful tools for diagnostics or therapeutic decision making\citep{woo2017building, wager2013fmri, spisak2020pain}. However, the presence of confounders typical for biomedical research (e.g. sample demographics, center-effects) and specific to the data acquisition and processing approach (e.g. imaging artifacts) presents a great challenge to these efforts.
The usefulness of the proposed tests is demonstrated in two such examples, using the HCP\citep{van2013wu}  and the ABIDE\citep{di2014autism} datasets.

In case of the HCP dataset, the weak but statistically significant confounder effect of age on fluid intelligence is in line with previous findings\citep{dubois2018distributed, lohmann2021predicting} and could likely exaggerate to a serious bias when testing the model on data of participants outside of the - relatively narrow - age range of the HCP sample. In this case, the bias would likely significantly harm the out-of-sample generalizability of this model. The bias of the same model for acquisition batch can also be problematic, especially as it has not yet been thoroughly discussed in case of the HCP dataset. There can be manifold reasons for the observed bias. Fluid intelligence of the included participants might be, for instance, affected by a changing selection bias during participant recruitment (e.g. as a consequence of the human connectome project receiving an increasing degree of public interest during its course).

In the ABIDE dataset, neither the center-bias nor the age-bias is surprising but both are obviously problematic for a diagnostic model of ASD. For instance the model trained on the raw (unadjusted) features - depending on the calibration of the predicted class probabilities - might classify all participants form e.g. the CMU (Carnegie Mellon University) center as neurotypical control participants. Similarly, the models biased by motion - next to having questionable neuroscientific validity - might systematically fail in populations with a tendency for higher in-scanner motion (as known for many conditions, among others ADHD\citep{eloyan2012automated} or Alzheimer's disease \citep{rao2017predictive}).

@@

On the other hand, the partial confounder test also showed, that the performance of the investigated confound mitigation techniques is much less convincing in the ABIDE dataset. The center bias of the classification in this massively multi-center dataset was found to be highly significant and, although mitigated, but not successfully removed by feature regression and COMBAT. Determining the relevance of the remaining bias is out of the scope of this paper, however significant p-values of the partial confounder test generally suggest that these models require more effective data harmonization.

Motion-bias in the ABIDE dataset was not detectable anymore after COMBAT, but remained highly significant after feature regression.
Interestingly, in this latter case, the association between the confounder and the predictions become even stronger than before but with an opposite sign. At the same time, the prediction performance substantially dropped (although still remained statistically significant). 
A possible explanation for this phenomenon is that in-scanner motion, as previously described\citep{fournier2010motor, anzulewicz2016toward}, has manifold links to ASD and, therefore, regressing out motion estimates from the connectivity features might eliminate a significant amount of signal-of-interest. This effect might be boosted by the previously reported 'residual motion artifacts' and their regional interactions\citep{spisak2014voxel} and raises caution for using feature regression to mitigate the effect of similar, complexly distributed confounders.

@@

 In sum, the application of the \emph{partial} confounder test on the real data examples suggests that confounding bias should be much more carefully investigated and reported in studies utilizing predictive modelling and machine learning as (i) variables as trivial as the date of the acquisition can cause significant model-bias and (ii) in certain situations, a sufficient mitigation of confounder-bias requires more effective solutions than feature regression and COMBAT and (iii) in some cases confounder mitigation can - paradoxically - introduce more bias. The partial confounder test can be considered as a useful, objective benchmark to guide the search for a suitable confounder mitigation approach for every dataset.
 Regarding the \emph{full} confounder test; while the real data examples in this paper were not typical cases of its potential applications, it has still provided important insights into model biases. Namely, in the case of the aforementioned paradox motion-regression model in the ABIDE dataset, the full confounder test did not provide any evidence that the model captures any extra variance over that explained by the confounder ($p=0.1$). In other words, the test accepted the null hypothesis of full model bias, suggesting that this model might be, in fact, almost exclusively driven by motion artifacts.
This example highlights that the full confounder test might become useful in exploratory phases of model development, where models might be still severely biased by various confounders and the questions is whether there is any biomedically relevant signal captured by the model, at all.

@@

To fill this critical gap in predictive model development, here I proposed two novel tests, the \emph{partial} and the \emph{full confounder tests}, which probe the null hypotheses of 'no confounder-bias' and 'full confounder-bias', respectively. 
The tests are distinguished from alternative approaches by their robustness to non-normally and non-linearly dependent predictions, rendering them applicable with a wide variety of machine learning models.
The tests have, moreover, a minimal computational overhead, as re-fitting the model is not required.

As demonstrated on functional brain connectivity-based predictive models of fluid intelligence and autism spectrum disorder, the tests can guide the optimization of the confound mitigation strategy and allow quantitative statistical assessment of the robustness, generalizability and neurobiological validity of predictive models in biomedical research.
Given their simplicity, robustness, wide applicability, high statistical power and computationally effective implementation (available in the  python package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}}), the partial and full confounder tests are novel tools in the methodological arsenal of predictive model development and might largely accelerate the development of clinically useful machine learning biomarkers.

@@

Obviously, a strong association between $\yhat$ and $\c$ might indicate that the model is biased; its predictions are driven by the confounder rather than information about the target variable.
Testing unconditional independence ($\yhat \independent \c$) between $\yhat$ and $\c$ (or any of the $\y$, $\yhat$, $\c$ variables) is, however, insufficient for the proper characterization of confounder-bias in predictive modelling.
For instance, even if $\yhat \independent \c$ is false, $\yhat$ might be only marginally dependent on $\c$, due to the dependence of both on $\y$. In other words, if the target variable $\y$ displays a true association to the confounder variable $\c$, a model that is completely blind to $\c$ (i.e not confounded at all) might still provide outputs $\yhat$ that are significantly associated with $\c$.

@@

Instead of focusing on the unconditional independence between the confounder and the predictions, we shall consider the \emph{conditional independence} between $\yhat$ and $\c$ given $\y$ (written as $\yhat \independent \c | \y$) which, by definition \citep{dawid1979conditional}, means that $\mathbb{P}(\yhat, \c|\y) = \mathbb{P}(\yhat|\y)\mathbb{P}(\c|\y)$. The statistical test with this null hypothesis will be referred to as the \emph{partial confounder test}.

@@

This result stands in strong contrast to \emph{unconditional} independence testing - where permutation tests \citep{pitman1937significance, fisher1942189}, provide a general, distribution-free solution - and it has important implications for confounder testing in predictive modelling where the distribution of the model outputs (conditioned on the target variable) - depending on the applied machine learning model - is unknown and often non-normal and non-linear.

One of the trivial candidates for the task, partial correlation, for instance assumes that all involved variables are multivariate Gaussian and - as to be shown below - even its Spearman-based variant is unable to tolerate relatively small deviations from normality and linearity.

@@

In the following section, CPT is adapted for partial confounder testing (option 3) and extended with general additive model\cite{hastie1987generalized} (GAM) and multinomial logistic regression\citep{bennett1966multiple, jones1975proability} based conditional distribution estimations, in order to make it handle categorical data and non-linear dependencies between the confounder and the target variable. (For an overview of the method, see Fig. \ref{fig:overview}). The formulation of the full confounder test (option 1) is analogous and given in Supplementary Material \ref{sup:full-test}.

@@

Smoothness of the GAM model is optimized with a grid-search by picking the model with the lowest generalized cross-validation score form the models defined by the default parameters as implemented in PyGAM\citep{serven2018generalized}.

@@

In the case of categorical $c$, a multinomial logistic regression model can be used to obtain $D(\cdot|\y)$, with the extra assumption of \emph{complete separation} if $y$ is also categorical (for an invertible Hessian, see e.g.\citep{bennett1966multiple, jones1975proability}).

Importantly, both the GAM-based and the multinomial logistic regression approaches guarantee that the numerator of Eq. \ref{eq-pperm} is always greater than zero and the group property for the permutations holds.

Note that from the three options for conditional independence-based null hypotheses enumerated in Table \ref{tab:conditional-independence-cases}, for option 2, the proposed approach can not provide a test that is assumption-free about $\yhat$, as the variable, on which the independence is conditional, must be always the predictor variable in Eq. \ref{eq:gam}. However, as discussed above, this option is of low practical relevance, anyway.
Pleasingly,  the proposed Gaussian regression-based conditional likelihood estimation ensures that no assumptions on $\yhat$ have to be made for the practically relevant options 1 and 3, i.e. for the full and partial confounder tests.

@@

Using CPT to test confounder-bias in predictive modelling allows relaxing assumptions on $\yhat$ (for both the partial and the full test) but - in line with the "no free lunch" theorem, requires knowing - or putting assumptions on - the joint distribution of the other two variables ($\y$ and $\c$). 
Berrett et al. \cite{berrett2020conditional} give a detailed analysis of the robustness of their CPT approach when estimating the conditional distribution with re-using the tested data via linear regression and, also, against misspecifying the conditional distribution of interest to introduce non-linearity.

Here I extend these results by performing simulations that evaluate the GAM- and multinomial liner model-based approaches, in a form that is accessible for power calculations in predictive modelling (considering various weights of the target signal in $\c$ and the confounder and the target signals in $\yhat$).
Moreover, I investigate the robustness of the tests against the violation of normality and linearity of the conditional distributions $D(\c|\y)$ and $D(\yhat|\y)$.

@@

The simulated predicted values are constructed in a similar fashion, but can depend on $\c$ as well:

@@

Note that simulations with $w_{c\hat{y}}=0$ produce data under the null hypothesis of no confounder bias.

>>>>

Estimates of statistical power for the partial confounder test (with normal and linear simulations, for  a wide range of parameters) are shown on Figure \ref{fig:sim-normal}. Notably, with sample sizes as large as 1000, a confounder contributing only $\sim 4\%$ to the variance of the predictions  ($w_{c\hat{y}} = 0.2$) can already be robustly detected with a power of 94-100\%. With a sample size of 500, the same confounding bias is still detected with a power greater than 84-100\% in all of the simulation cases. A sample size of 100 requires a somewhat stronger bias with approximately 12\% of explained variance ($w_{c\hat{y}}=0.4$) to achieve a reasonable level of power (75-98\%). Finally, even with a relatively low sample size of 50, the same amount of confounder variance is detected with a power of at least 50\%. If the confounder explains more than 25\% of variance, it is almost certainly detected even with a low sample size of $n \geq 50$.

Simulations show that non-normality has minimal effect on the power of the tests, except in case of extreme non-normality. (Fig. \ref{fig:sim-non-normal}). Simulations with sigmoid dependence resulted in an apparent loss of statistical power, however this is simply a consequence of the simulation methodology: with the same parameters, the sigmoid transformed confounder explains only approximately half the variance as compared to linear simulations.
Type I error control was valid in case of categorical variables and for the \emph{full} confounder test, as well. Power in these cases was also highly similar to the that of the partial confounder test with numerical variables. (Supplementary Figures \ref{fig:sim-bbb-lin-partial}, \ref{fig:sim-bbb-lin-full}, \ref{fig:sim-bbb-sig-partial} and \ref{fig:sim-bbb-sig-full}).

@@

Both acquisition batch and age group were statistically significantly associated with $G_f$ ($R^2=0.032$ and $0.011$ and $p<0.001$ and $p=0.001$, respectively, see also Table \ref{tab:unconditional-pvals}). The model trained on the raw (unadjusted) connectivity features predicted fluid intelligence with a medium effect size ($R^2=0.095$, $p<0.001$).

@@

The full confounder test was highly significant ($p<0.0001$) for all models, indicating that neither of the models were exclusively driven by the confounders.

@@

The full confounder test was strongly significant ($p<0.0001$) for all models, except when regressing-out motion from the features ($p=0.09$), indicating that this model may be (almost) exclusively driven by motion artifacts.

@@

\end{tabular}
\caption{\label{tab:unconditional-pvals} Coefficients-of-determination ($R^2$), the corresponding p-values and the p-values of the partial and full confounder tests, for all investigated datasets, confounders (conf.) and confounder-mitigation methods (method). Bold numbers denote significant confounding bias identified by the partial confounder test or the lack of significant contribution over that of the confounder, as identified by the full confounder test.}
\end{table}

@@

The assessment of confounding bias via conditional independence testing is a very appealing solution due to its simplicity and applicability in a wide variety of settings. However, statistical testing of conditional independence is challenging: 'conditional analogs' of bivariate (unconditioned) non-parametric tests can give invalid results even by slight violations of normality and linearity\citep{korn1984ranges} (see an example on Fig. \ref{fig:sim-h0-demo}). While the magnitude of this problem may not be fully appreciated in case of predictive model diagnostics, the 'no free lunch theorem' (i.e. the nonexistence of a uniformly valid conditional independence test)\cite{shah2020hardness} clearly implies that conditional independence-based confounding bias testing must be designed so that its suitability for the particular problem may be judged easily.

@@

As expected from theory, both tests displayed a valid type I error control and a high, practically relevant statistical power in the simulations, even if both the predictions and the confounder are non-normally and/or non-linearly dependent on the target variable (except by extreme non-normality). This result confirms that the tests can be deployed in a wide variety of predictive modelling scenarios. While different biomedical applications might consider different amounts of bias to be relevant, the presented results can serve as a basis for power calculations, in order to identify the necessary sample size for proper model diagnostics.

A characteristic example for the potential areas of applications is the novel field of "predictive neuroscience", where applying predictive modelling and machine learning on functional neuroimaging data holds great potential for both revolutionizing our understanding of the physical basis of mind and delivering clinically useful tools for diagnostics or therapeutic decision making\citep{woo2017building, wager2013fmri, spisak2020pain}. However, the presence of confounders that are typical for biomedical research (e.g. sample demographics, center-effects) or specific to the data acquisition and processing approach (e.g. imaging artifacts) presents a great challenge to these efforts.
The usefulness of the proposed tests is demonstrated in two such examples, using the HCP\citep{van2013wu}  and the ABIDE\citep{di2014autism} datasets.

In case of the HCP dataset, the statistically significant age bias of the 'raw' model for predicting fluid intelligence is in line with previous findings\citep{dubois2018distributed, lohmann2021predicting} and could likely exaggerate to a serious bias when testing the model on data of participants outside of the - relatively narrow - age range of the HCP sample. In this case, the bias would likely significantly harm the out-of-sample generalizability of this model. The bias of the same model for acquisition batch can also be problematic, especially as it has not yet been thoroughly discussed in case of the HCP dataset. There can be manifold reasons for the observed acquisition bias. Fluid intelligence of the included participants might be, for instance, affected by a changing selection bias during participant recruitment (e.g. as a consequence of the human connectome project receiving an increasing degree of public interest during its course).

In the ABIDE dataset, neither the center bias nor the age bias is surprising in the case of the 'raw' model but both would be obviously severely problematic for a diagnostic biomarker candidate of ASD. For instance the model trained on the raw (unadjusted) features - depending on the calibration of the predicted class probabilities - might classify all participants form e.g. the CMU (Carnegie Mellon University) center as neurotypical control participants. Similarly, the models biased by motion - next to having questionable neuroscientific validity - might systematically fail in populations with a tendency for higher in-scanner motion (as known for many conditions, among others ADHD\citep{eloyan2012automated} or Alzheimer's disease \citep{rao2017predictive}).

@@

On the other hand, the partial confounder test also showed, that the performance of the investigated confound mitigation techniques is much less convincing in the ABIDE dataset. The center bias of the classification in this massively multi-center dataset was, although mitigated, but not successfully removed by feature regression and COMBAT. Determining the relevance of the remaining bias is out of the scope of this paper, however significant p-values of the partial confounder test generally suggest that these models require more effective data harmonization.

Motion bias in the ABIDE dataset was not detectable anymore after COMBAT, but remained highly significant after feature regression.
Interestingly, in this latter case, the association between the confounder and the predictions become even stronger than before but with an opposite sign. At the same time, the prediction performance substantially dropped (although still remained statistically significant). 
A possible explanation for this phenomenon is that in-scanner motion, as previously described\citep{fournier2010motor, anzulewicz2016toward}, has manifold links to ASD and, therefore, regressing out motion estimates from the connectivity features might eliminate a significant amount of signal-of-interest. This effect might be boosted by the previously reported 'residual motion artifacts' and their regional interactions\citep{spisak2014voxel} and raises caution for using feature regression to mitigate the effect of similar, complexly distributed confounders.

@@

 In sum, the application of the \emph{partial} confounder test on the real data examples suggests that confounding bias should be much more carefully investigated and reported in studies utilizing predictive modelling and machine learning as (i) variables as trivial as the date of the acquisition can cause significant confounding bias and (ii) in certain situations, a sufficient mitigation of confounder-bias requires more effective solutions than feature regression or COMBAT and (iii) in some cases confounder mitigation can - paradoxically - introduce more bias. The partial confounder test can be considered as a useful, objective benchmark to guide the search for a suitable confounder mitigation approach for every dataset.

 Regarding the \emph{full} confounder test; while the real data examples in this paper were not typical cases of its potential applications, it has still provided important insights into model biases. Namely, in the case of the aforementioned paradox motion-regression model in the ABIDE dataset, the full confounder test did not provide any evidence that the model captures any extra variance over that explained by the confounder ($p=0.1$). In other words, the test accepted the null hypothesis of full model bias, suggesting that this model might be, in fact, almost exclusively driven by motion artifacts.
This example highlights that the full confounder test might become useful in exploratory phases of model development, where models might be still severely biased by various confounders and the questions is whether there is any biomedically relevant signal captured by the model, at all.

@@

To fill this critical gap in predictive model development, here I proposed two novel tests, the \emph{partial} and the \emph{full confounder tests}, which probe the null hypotheses of 'no confounding bias' and 'full confounding bias', respectively. 
The tests are distinguished from alternative approaches by their robustness to non-normally and non-linearly dependent predictions, rendering them applicable with a wide variety of machine learning models.
The tests have, moreover, a minimal computational overhead, as re-fitting the model is not required.

As demonstrated on functional brain connectivity-based predictive models of fluid intelligence and autism spectrum disorder, the tests can guide the optimization of the confound mitigation strategy and allow quantitative statistical assessment of the robustness, generalizability and neurobiological validity of predictive models in biomedical research.
Given their simplicity, robustness, wide applicability, high statistical power and computationally effective implementation (available in the  python package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}}), the partial and full confounder tests emerge as novel tools in the methodological arsenal of predictive modelling and might largely accelerate the development of clinically useful machine learning biomarkers.

@@

Obviously, a strong association between $\yhat$ and $\c$ might indicate that the model is biased; its predictions are driven by the confounder rather than information about the target variable.
Assessing the simple bivariate (unconditioned) dependence ($H0: \yhat \independent \c$) between $\yhat$ and $\c$ (or any of the $\y$, $\yhat$, $\c$ variables) is, however, insufficient for the proper characterization of confounder-bias in predictive modelling.
For instance, even if $\yhat \independent \c$ is false, $\yhat$ might be only marginally dependent on $\c$, due to the dependence of both on $\y$. In other words, if the target variable $\y$ displays a true association to the confounder variable $\c$, a model that is completely blind to $\c$ (i.e not confounded at all) might still provide outputs $\yhat$ that are significantly associated with $\c$.

@@

Instead of focusing on the 'unconditioned' independence between the confounder and the predictions, we shall consider the \emph{conditional independence} between $\yhat$ and $\c$ given $\y$ (written as $\yhat \independent \c | \y$) which, by definition \citep{dawid1979conditional}, means that $\mathbb{P}(\yhat, \c|\y) = \mathbb{P}(\yhat|\y)\mathbb{P}(\c|\y)$. The statistical test with this null hypothesis will be referred to as the \emph{partial confounder test}.

@@

This result stands in strong contrast to \emph{unconditional} independence testing - where permutation tests \citep{pitman1937significance, fisher1942189}, provide a general, distribution-free solution - and it has important implications for confounder testing in predictive modelling where the distribution of the model outputs (conditioned on the target variable) - depending on the applied machine learning model - is unknown and often non-normal and non-linear. One of the trivial candidates for the task, partial correlation, for instance assumes that all involved variables are multivariate Gaussian and - as to be shown below in a simulated example - even its Spearman-based variant is unable to tolerate relatively small deviations from normality and linearity.

@@

In the following section, CPT is adapted for \emph{partial} confounder testing (option 3) and extended with general additive model\cite{hastie1987generalized} (GAM) and multinomial logistic regression\citep{bennett1966multiple, jones1975proability} based conditional distribution estimations, in order to make it handle categorical data and non-linear dependencies between the confounder and the target variable. (For an overview of the method, see Fig. \ref{fig:overview}). The formulation of the \emph{full} confounder test (option 1) is analogous and given in Supplementary Material \ref{sup:full-test}.

@@

Smoothness of the GAM model is optimized with a grid-search by picking the model with the lowest generalized cross-validation score form the models defined by the default parameters as implemented in PyGAM\citep{serven2018generalized} (v0.8.0).

@@

In the case of categorical $\c$, a multinomial logistic regression (\emph{mnlogit}) model can be used to obtain $D(\cdot|\y)$, with the extra assumption of \emph{complete separation} if $\y$ is also categorical (in order to ensure an invertible Hessian, see e.g. Ref.\citep{bennett1966multiple, jones1975proability}).

Importantly, both the GAM- and the \emph{mnlogit}-based approaches guarantee that the numerator of Eq. \ref{eq-pperm} is always greater than zero and the group property for the permutations holds.

Note that from the three options for conditional independence-based null hypotheses enumerated in Table \ref{tab:conditional-independence-cases}, the proposed approach can not provide a test for option 2 that is assumption-free about $\yhat$, as the variable, on which the independence is conditional, must be always the predictor variable in Eq. \ref{eq:gam}. However, as discussed above, this option is of low practical relevance, anyway.
Pleasingly,  the proposed Gaussian regression-based conditional likelihood estimation ensures that no assumptions on $\yhat$ have to be made for the practically relevant options 1 and 3, i.e. for the full and partial confounder tests.

@@

Using CPT to test confounding bias in predictive modelling allows relaxing assumptions on $\yhat$ (for both the partial and the full test) but - in line with the "no free lunch" theorem, requires knowing - or putting assumptions on - the joint distribution of the other two variables ($\y$ and $\c$). 
Berrett et al. \cite{berrett2020conditional} give a detailed analysis of the robustness of their CPT approach when estimating the conditional distribution with re-using the tested data via linear regression and, also, against misspecifying the conditional distribution of interest to introduce non-linearity.

Here I extend these results by performing simulations that evaluate the GAM- and \emph{mnlogit}-based approaches, in a form that is accessible for power calculations in predictive modelling (considering various weights of the target signal in $\c$ and the confounder and the target signals in $\yhat$).
Moreover, I investigate the robustness of the tests against the violation of normality and linearity of the conditional distributions $D(\c|\y)$ and $D(\yhat|\y)$.

@@

The simulated predicted values are constructed in a similar fashion, but may depend on $\c$ as well:

@@

Note that simulations with $w_{c\hat{y}}=0$ produce data under the null hypothesis of no confounding bias.
